Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-18 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex

We have added some text to section II (between (14)-(15)) and a sentence to section IV (between \todo{(26)-(27)}) to clarify these issues.

\begin{referee}
Here the conservation properties of mass and energy in three trajectory models are discussed,
but how about the parallel momentum balance? Relating to this question, in page 20 - 21
authors have discussed about the relation between the momentum balance and friction force
from different collision operator models. I agree with the authors that the momentum
conservation property in collision operator is important to evaluate correct transport coefficients.
However, by seeing the figures 3(c)(d), parallel momentum balance seems to be broken in the
partial trajectory model, even if the full linearized Fokker-Planck operator is used.
The momentum balance equation (36) is derived by taking the $\vect{b}\cdot \int d^3v\, m\vect{v}$ moment of the drift
kinetic equation (15). Then, for each three trajectory model, taking the $\vect{b}\cdot \int d^3v\, m\vect{v}$ moment of
eq. (15) with combining one of eq. (17),(18), or (19), does it yield the proper parallel momentum
balance equation as eq.(36)? Or more simply, do these three trajectory models evaluate the
parallel viscosity $\left<\vect{b}\cdot\nabla\cdot\vect{\pi}\right>$ correctly?
\end{referee}

We have added a new section IV to the paper to discuss momentum balance.
To summarize, we first discuss how parallel momentum has a different status to mass and energy in that mass and
energy are conserved, whereas parallel momentum is not conserved due to the mirror force.
We then compare the $m v_{||}$ moments of the various forms of the drift-kinetic
equation to the full fluid momentum equation, as you suggest.
We show the electric field terms in the kinetic perspective
are associated with a piece of the gyroviscosity in the fluid perspective.  The full trajectory model
gives the correct viscous force whereas the other trajectory models do not.
One further noteworthy property of the parallel velocity moments of the kinetic equations
is that of the trajectory models considered, only the full trajectories preserve intrinsic
ambipolarity in quasisymmetry when $E_r \ne 0$. This result is now discussed at the end of section IV.

@@

We agree that in scenarios with $T_e \gg T_i$, such as a Core Electron Root Confinement regime,
$E_*$ may be larger than a few percent. (Our original statement
was meant only to apply to a standard ion-root W7-X regime,
but this should have been clarified.) The text in the manuscript has now been updated to acknowledge
the possibility of an electron root and that $E_*$ may not be small.

\begin{referee}
Appendix: The authors claims that the parallel flow and heat flux solved by SFINCS code obeys the
isomorphism, but is radial electric field ($d\Phi/dr$) contained in ``other input quantities''? Please
clarity that. Also, if the isomorphism is satisfied for finite $d\Phi/dr$ and finite source term, I
suppose that the usage of source term in the full trajectory model is justified at least in
quasisymmetric systems. Can author comment on this point in Appendix?
\end{referee}

We have changed and expanded the text in this appendix for clarification. 
For all of the trajectory models described, an isomorphism does hold for the transport matrix elements
if the electric field is varied so
$(GM+IN)(\iota M-N)^{-1} d\Phi_0/d\psi$ remains constant as $M$ and $N$ are varied.
%This required transformation of $d\Phi_0/d\psi$ can be found in the last line of table 1 in Ref. [49].

The second part of your question is an interesting observation: when the full trajectory model
is applied to a quasisymmetric field, the source vanishes for any $E_r$, not just at a special ambipolar $E_r$ (as in a non-quasisymmetric field.)
We now state this fact in the new discussion of intrinsic ambipolarity at the end of section IV.

### 2014-02-18 doc/sfincsPaper/sfincsPaper.tex
Added discussion of momentum balance to revised manuscript and referee response

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}
%\newcommand{\changed}[1]{#1}
\newcommand{\changed}[1]{\textcolor{red}{#1}}

@@

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
\end{equation}
where $\vect{r}$ denotes the position vector,
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
for several reasons.
First, if the  $\vma$ term in (\ref{eq:rdot1}) were retained, we would need to solve a 5D rather
than 4D problem due to the radial coupling (i.e. $\psi$ appearing
as a derivative rather than merely as a parameter). Second,
once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
in order for $\mu$ to be conserved.
Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
on $d T_a/d\psi$, $dB/d\psi$,
\changed{
and $\rho_*$.
Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
(while keeping $d\Phi_0/d\psi$ finite), 
and this limit is already complicated and interesting to explore without the extra complexity
of finite-$\rho_*$ corrections.
Fifth, 
we wish to focus on the effects of the radial electric field.
The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
Finally,
these omitted terms would significantly complicate the analysis in section III,
in which we will examine moments of the kinetic equation.
}

@@

\section{Conservation properties and sources}
\label{sec:sources}

If one attempts to solve the kinetic equation (\ref{eq:kinetic4}) numerically using either the full or partial trajectories
and $E_r \neq 0$, unphysical results will be obtained, with
the numerical solution not converging as resolution parameters are increased.
We now explore the reason for this behavior.
We will then describe a modified form of the kinetic equation which robustly produces more sensible results.

@@

@@

When $\nu' < 1$, the resolution required in the $\theta$, $\zeta$, and $\xi$ coordinates
increases as $\nu'$ decreases,
due to the boundary layers that develop in phase space.  The highest resolution used
for results presented here, corresponding to the W7-X calculations at  $\nu'=10^{-3}$,
was $N_\theta=29$, $N_\zeta = 83$, $N_\xi=180$, and $N_x=5$,
giving a $2,166,302 \times 2,166,302$ linear system.
Here, $N_j$ is the number of grid points or modes in coordinate $j$.
Each calculation at this resolution with the Fokker-Planck collision operator
required $\sim 30-50$ minutes to run on 4 nodes of the Edison computer at NERSC.
At higher collisionality, lower resolution is sufficient, so memory and time requirements are reduced;
computations with $\nu' > 10^{-2}$ can typically be run on a laptop.

@@

A useful test of a stellarator neoclassical code such as the one
described here is the quasisymmetry isomorphism,
discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
\changed{
A quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
for some periodic function $y$ and integers $M$ and $N$.
Magnetic fields with the same $y$ but different $M$ and $N$ are said to be isomorphic in that
the associated transport matrices must be related in the following manner.
Suppose the transport matrices are computed for several quasisymmetric magnetic fields with different values of $M$ and $N$,
varying the collision frequency in each calculation so
$\nuii / (\iota M - N)$ remains fixed,
and varying the radial electric field so
$(d\Phi_0/d\psi)(GM+IN)/(\iota M-N)$
remains fixed.
%Other input parameters such as $dn/d\psi$ and $dT/d\psi$ are kept fixed
%as $M$ and $N$ are varied.
%Under these circumstances, 
In such a scan of $M$ and $N$,
it can be shown analytically \cite{Pytte, Boozer83, meQS} that the 
transport matrix elements should vary as follows:
$L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$ vary
$\propto (NI+MG)^2/(\iota M-N)$;
$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ vary
$\propto (NI+MG)/(\iota M-N)$;
and $L_{33}$ varies
$\propto 1/(\iota M-N)$.
This isomorphism holds for all the trajectory models considered in this paper.
}
\begin{comment}
flow
and heat flux should obey the transformations
\begin{equation}
\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
\end{equation}
and
\begin{equation}
\left< \int d^3v\, f_{\mathrm{i}1} \frac{v^2}{2}\vect{v}_{\mathrm{m}i}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
\end{equation}
respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
\end{comment}

@@

\changed{
Expressions for the matrix coefficients in the Pfirsch-Schl{\"u}ter regime are summarized in 
Eqs.~(\ref{eq:LPSregime})-(\ref{eq:PSregimeFunctions}). 
Three numerical coefficients in the function ${K}_2^{\mathrm{Simakov}} \left(\psi\right)$
arise from generalized Spitzer problems, 
which were solved in Ref.~\cite{AndreiPer2009} by keeping a small number of Laguerre polynomials in kinetic energy.
When these generalized Spitzer problems are solved numerically keeping many more energy polynomials,
we obtain the more accurate coefficients given in ${K}_{2}$ below.
}

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\tens}[1]{\overleftrightarrow{#1}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}
%\newcommand{\changed}[1]{#1}
\newcommand{\changed}[1]{\textcolor{red}{#1}}

@@

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
\changed{In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
and demonstrate that the electric field terms in the kinetic description
correspond to a component of gyroviscosity in a fluid description.}
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
\end{equation}
where $\vect{r}$ denotes the position vector,
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
for several reasons.
First, if the  $\vma$ term in (\ref{eq:rdot1}) were retained, we would need to solve a 5D rather
than 4D problem due to the radial coupling (i.e. $\psi$ appearing
as a derivative rather than merely as a parameter). Second,
once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
in order for $\mu$ to be conserved.
Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
on $d T_a/d\psi$, $dB/d\psi$,
\changed{
and $\rho_*$.
Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
(while keeping the $d\Phi_0/d\psi$ terms finite), 
and this limit is already complicated and interesting to explore without the extra complexity
of finite-$\rho_*$ corrections.
Fifth, 
we wish to focus on the effects of the radial electric field.
The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
Finally,
these omitted terms would significantly complicate the analysis in section \ref{sec:sources},
in which we will examine moments of the kinetic equation.
}

@@

\section{Particle and energy moment equations, conservation properties, and sources}
\label{sec:sources}

If one attempts to solve the kinetic equation (\ref{eq:kinetic4}) numerically using either the full or partial trajectories
and $E_r \neq 0$, unphysical results will be obtained, with
the numerical solution not converging as resolution parameters are increased.
We now explore the reason for this behavior.
We will then describe a modified form of the kinetic equation which robustly produces more sensible results.
\changed{The issues discussed in this section are related to moment equations for mass and energy; momentum has a different status
and will be examined in the next section.}

@@

\section{Momentum moment equations}
\label{sec:momentum}

\changed{
Parallel momentum has a different status to density and energy, in that density and energy are conserved by
the collisionless motion while parallel momentum is not, due to the mirror force.
(For example, considering the simple case of a single ion species with $d\Phi_0/d\psi=0$, $F_i$ and $v^2 F_i$ are homogeneous
solutions to the kinetic equation, whereas $v_{||} F_i$ is not.)
A consequence is that there does not appear to be a false constraint for $E_r \ne 0$ arising
from the $\left< \int d^3 v\, v_{||} (\ldots)\right>$ moment of the various forms of the kinetic equation,
i.e. there is no exact analogue to (\ref{eq:unphysicalMassConservation}), (\ref{eq:unphysicalEnergyConservation1}),
or (\ref{eq:unphysicalEnergyConservation2}) for momentum.
When the momentum moment of the various forms of the drift-kinetic equation is taken,
even if a factor of $B$ or $1/B$ is included in the flux surface average, 
a collisionless term remains that is not proportional to $d\Phi_0/d\psi$.
Consequently, for all the trajectory models,
$d\Phi_0/d\psi=0$ is a well behaved rather than singular
limit of the momentum moment equation.
}

\changed{
Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
for each trajectory model to the full parallel momentum fluid equation. This later equation is
\begin{equation}
???
\end{equation}
First, consider the case of no radial electric field.
The stress tensor is given to the accuracy needed by $\tens{\Pi}_a \approx \tens{\Pi}_{a||}$
where $\tens{\Pi}_{a||} = p_{a \bot} \tens{I} + (p_{a||} - p_{a\bot}) \vect{b}\vect{b}$,
$p_{a||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
Then noting the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ v_{||} \vect{b}\cdot\nabla f_{a1} - \frac{1-\xi^2}{2B}v(\vect{b}\cdot\nabla B) \frac{\partial f_{a1}}{\partial \xi} \right]
& = &
\vect{b}\cdot\nabla p_{a||1} + \frac{ p_{a\bot 1} - p_{a || 1}}{B} \vect{b}\cdot\nabla B \nonumber \\
& = &
 -\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a||} \right),
\end{eqnarray}
one can see the drift-kinetic equation (\ref{}) is consistent with the full fluid parallel momentum equation
at least when $E_r=0$.
}

\changed{
Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = -\frac{c}{B} \vect{v}\cdot\vect{b}\times\nabla\Phi \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different indepedent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, \bar{f}_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\Phi \approx \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}
}

\changed{
Finally, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
terms in the drift-kinetic equation for various trajectory models, to see if the results agree with (\ref{eq:PiEParallel}).
For the full trajectories, the moment of the $d\Phi_0/d\psi$ terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{B^2} \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
- (\vma \cdot\nabla\psi) \frac{Z_a e}{2 T_a x_a} \frac{d\Phi_0}{d\psi} \frac{\partial f_{a1}}{\partial x_a}
\right.
\nonumber \\
\left.
+ \xi(1-\xi^2) \frac{c}{2B^3} \frac{d\Phi_0}{d\psi} (\vect{B}\times\nabla\psi\cdot\nabla B)\frac{\partial f_{a}}{\partial \xi}
\right]
\nonumber \\
= 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right),
\label{eq:fullViscosity}
\end{eqnarray}
obtained by integrating by parts in $x_a$ and $\xi$.
Thus, the full-trajectory model agrees with the full fluid parallel momentum equation: (\ref{eq:fullViscosity}) = (\ref{eq:PiEParallel}).
However, this agreement is
not shared by the DKES model: the moment of the $d\Phi_0/d\psi$ term in (\ref{eq:DKESTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{\left< B^2 \right> } \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
\right]
= 
\frac{c m_a}{ \left< B^2 \right>}
\frac{d\Phi_0}{d\psi}
\vect{B}\times\nabla\psi\cdot\nabla (n_a V_{a||}),
\label{eq:DKESViscosity}
\end{eqnarray}
which does not equal (\ref{eq:PiEParallel}). The corresponding result for the partial trajectories,
obtained by replacing $\left< B^2 \right> \to B^2 $ in (\ref{eq:DKESViscosity}), also does not match
(\ref{eq:PiEParallel}). Thus, the DKES and partial trajectory models do not correctly account for the parallel
viscous force as the full trajectory model does.
}

\changed{
We close this section by noting another an important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, the $d\Phi_0/d\psi$ terms
may be integrated by parts, canceling each other to leave
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning the radial current vanishes. However, for the DKES and partial trajectory models, the $d\Phi_0/d\psi$ terms
do not cancel, leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
meaning the radial current is generally nonzero.  Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

When $\nu' < 1$, the resolution required in the $\theta$, $\zeta$, and $\xi$ coordinates
increases as $\nu'$ decreases,
due to the boundary layers that develop in phase space.  The highest resolution used
for results presented here, corresponding to the W7-X calculations at  $\nu'=10^{-3}$,
was $N_\theta=29$, $N_\zeta = 83$, $N_\xi=180$, and $N_x=5$,
giving a $2,166,302 \times 2,166,302$ linear system.
Here, $N_j$ is the number of grid points or modes in coordinate $j$.
Each calculation at this resolution with the Fokker-Planck collision operator
required $\sim 30-50$ minutes to run on 4 nodes of the Edison computer at NERSC.
At higher collisionality, or if fewer harmonics are retained in $B(\theta,\zeta)$, 
lower resolution is sufficient, so memory and time requirements are reduced;
\changed{
for example, in the same W7-X geometry at $\nu'=10^{-2}$, sufficient resolution parameters for convergence
were $N_\theta=11$, $N_\zeta=64$, $N_\xi=100$, and $N_x=5$,
and computations required 3 minutes on 1 node of Edison.}
Computations with $\nu' > 10^{-2}$ can typically be run on a laptop.

@@

A useful test of a stellarator neoclassical code such as the one
described here is the quasisymmetry isomorphism,
discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
\changed{
Equivalent to the definition at the end of section \ref{sec:sources} \cite{PerAndreiPRL},
a quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
for some periodic function $y$ and integers $M$ and $N$.
Magnetic fields with the same $y$ but different $M$ and $N$ are said to be isomorphic in that
the associated transport matrices must be related in the following manner.
Suppose the transport matrices are computed for several quasisymmetric magnetic fields with different values of $M$ and $N$,
varying the collision frequency in each calculation so
$\nuii / (\iota M - N)$ remains fixed,
and varying the radial electric field so
$(d\Phi_0/d\psi)(GM+IN)/(\iota M-N)$
remains fixed.
%Other input parameters such as $dn/d\psi$ and $dT/d\psi$ are kept fixed
%as $M$ and $N$ are varied.
%Under these circumstances, 
In such a scan of $M$ and $N$,
it can be shown analytically \cite{Pytte, Boozer83, meQS} that the 
transport matrix elements should vary as follows:
$L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$ vary
$\propto (NI+MG)^2/(\iota M-N)$;
$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ vary
$\propto (NI+MG)/(\iota M-N)$;
and $L_{33}$ varies
$\propto 1/(\iota M-N)$.
This isomorphism holds for all the trajectory models considered in this paper.
}
\begin{comment}
flow
and heat flux should obey the transformations
\begin{equation}
\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
\end{equation}
and
\begin{equation}
\left< \int d^3v\, f_{\mathrm{i}1} \frac{v^2}{2}\vect{v}_{\mathrm{m}i}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
\end{equation}
respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
\end{comment}

@@

\changed{
Expressions for the matrix coefficients in the Pfirsch-Schl{\"u}ter regime are summarized in 
Eqs.~(\ref{eq:LPSregime})-(\ref{eq:PSregimeFunctions}). 
Three numerical coefficients in the function ${K}_2^{\mathrm{Simakov}} \left(\psi\right)$
arise from generalized Spitzer problems, 
which were solved in Ref.~\cite{AndreiPer2009} by keeping a small number of Laguerre polynomials in kinetic energy.
When these generalized Spitzer problems are solved keeping more energy polynomials,
we obtain the more accurate coefficients given in ${K}_{2}$ below.
}
